Efficiency

### 2018-01-02 pdfs/Untitled.Rmd

---
references:

- id: garland2005
  title: Elementary Physics of Reactor Control Module - Fission Product Poisoning
  author:
  - family: Garland
    given: Bill
  volume: 3
  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
  publisher: Nuclear Engineering (Originally from McMaster University)
  type: article-journal
  issued:
    year: 2005

- id: fujisawa1995
  title: Measurement of Basic Performances for Waterjet Propulsion Systems in Water Tunnel
  author:
  - family: Fujisawa
    given: Nobuyuki
  volume: 2
  URL: 'downloads.hindawi.com/journals/ijrm/1995/936068.pdf'
  publisher: International Journal of Rotating Machinery
  type: article-journal
  issued:
    year: 1995

- id: carlton2007
  title: Marine Propellers and Propulsion
  author:
  - family: Carlton
    given: John
  URL: 'http://www.nuceng.ca/br_space/2017-09_4d03_6d03/misc_files/xenon_poisoning.pdf'
  publisher: Elsevier
  issued:
    year: 2007

- id: liang2014
  title: The economics of slow steaming
  author:
  - family: Liang
    given: Lee Hong
  URL: 'http://www.seatrade-maritime.com/news/americas/the-economics-of-slow-steaming.html'
  publisher: Seatrade Maritime News
  issued:
    year: 2014
    month: 10
    day: 7

- id: sanguri2012
  title: The Guide to Slow Steaming On Ships
  author:
  - family: Sanguri
    given: Mohit
  URL: 'https://www.marineinsight.com/wp-content/uploads/2013/01/The-guide-to-slow-steaming-on-ships.pdf'
  publisher: Marine Insight
  issued:
    year: 2012
    month: 12

- id: WNA2017
  title: Nuclear Powered Ships
  author:
  - family: 
    given: 
  URL: 'http://www.world-nuclear.org/information-library/non-power-nuclear-applications/transport/nuclear-powered-ships.aspx#ECSArticleLink0'
  publisher: World Nuclear Association
  issued:
    year: 2017

- id: WNA2009
  title: Chernobyl Accident Appendix 1
  author:
  - family: 
    given: 
  URL: 'http://www.world-nuclear.org/information-library/safety-and-security/safety-of-plants/appendices/chernobyl-accident-appendix-1-sequence-of-events.aspx'
  publisher: World Nuclear Association
  issued:
    year: 2009

- id: NPdotnet2017
  title: Laminar Flow
  author:
  - family: 
    given: 
  URL: 'http://www.nuclear-power.net/nuclear-engineering/fluid-dynamics/laminar-flow-viscous/'
  publisher: nuclear-power.net
  issued:
    year: 2017

- id: patrick2012
  title: The Driving Factor in the SEA 1000 choice
  author:
  - family: Patrick
    given: Rex
  URL: 'https://corporate.siemens.com.au/content/dam/internet/siemens-com-au/root/aunz-defence-solutions/apdr-october-2015-issue-future-submarine.pdf'
  publisher: Asia-Pacific Defence Reporter
  issued:
    year: 2015
    month: 10

- id: patrick2017
  title: Senate Estimates Committee 
  author:
  - family: Patrick
    given: Rex
  URL: 'http://parlview.aph.gov.au/mediaPlayer.php?videoID=379261&operation_mode=parlview#/4%20from%20around%201201'
  publisher: Australian Parliament House
  issued:
    year: 2017
    month: 10

- id: hamilton1997
  title: Jet Torque
  author:
  - family: Waterjets
    given: Hamilton
  volume: 8
  URL: 'http://www.hamiltonmarine.co.nz/includes/files_cms/file/JetTorque%2008.pdf'
  publisher: Hamilton Waterjets
  issued:
    year: 1997
    month: 3

- id: mollard2011
  title: Ship Resistance and Propulsion
  author:
  - family: Mollard
    given: Anthony F.
  author:
  - family: Turnock
    given: Stephen R.
  author:
  - family: Hudson
    given: Dominic A.
  URL: 'http://dl.kashti.ir/ENBOOKS/NEW/Ship%20resistance%20and%20propulsion%20_%20F.Molland%20(2011).pdf'
  publisher: Cambridge University Press
  issued:
    year: 2011

- id: bruce1974
  title: The Design of Pumpjets for Hydrodynamic Propulsion
  author:
  - family: Bruce
    given: Edgar P.
  author:
  - family: Gearhart
    given: Walter S.
  author:
  - family: Ross
    given: John R.
  - family: Treaster
    given: Allen L.
  URL: 'https://ntrs.nasa.gov/archive/nasa/casi.ntrs.nasa.gov/19750003135.pdf'
  publisher: Ordnance Research Laboratory, The Pennsylvania State University
  issued:
    year: 1974
    month: 01

  - id: furuya1988
  title: A new pumpjet design theory
  author:
  - family: Furuya
    given: Oritsugu
  author:
  - family: Chiang
    given: Wen-Li
  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/a201353.pdf'
  publisher: Office of Naval Research
  issued:
    year: 1988
    month: 6

- id: shin2015
  title: Methods for Cavitation Prediction on Tip-Modified Propellers in Ship Wake Fields
  author:
  - family: Shin
    given: Keun Woo
  - family: Regener
    given: Pelle Bo
  - family: Andersen
    given: Poul
  URL: 'http://www.marinepropulsors.com/proceedings/2015/WA2-1.pdf'
  publisher: Fourth International Symposium on Marine Propulsors
  issued:
    year: 2015
    month: 6

- id: moulijn2015
  title: Application of Various Computational Methods for Predict the Performance of Cavitation of Ducted Propellers
  author:
  - family: Moulijn
    given: Joost
  URL: 'www.marinepropulsors.com/proceedings/2015/WB3-1.pdf'
  publisher: Fourth International Symposium on Marine Propulsors
  issued:
    year: 2015
    month: 6

- id: lewis1988
  title: Principals of Naval Architecture
  author:
  - family: Van Manen
    given: J.D.
  - family: Van Oossanen
    given: P.
  volume: 2
  URL: 'http://opac.vimaru.edu.vn/edata/EBook/Principles%20of%20Naval%20achitecture%202.pdf'
  publisher: The Society of Naval Architects and Marine Engineers
  editor: 
  - family: Lewis
    given: Edward V.
  issued:
    year: 1988

- id: li2013
  title: Dynamic Characteristics of Rotating Stall in Mixed Flow Pump
  author:
  - family: Li
    given: Xiaojun
  - family: Yuan
    given: Shouqi
  - family: Pan
    given: Zhongyong
  - family: Li
    given: Yi
  - family: Liu
    given: Wei
  URL: 'https://www.hindawi.com/journals/jam/2013/104629/'
  publisher: Hindawi Publishing Corporation Journal of Applied Mathematics
  issued:
    year: 2013
    month: 8

- id: kamal2015
  title: Powering for Medium Speed Wave-Piercing Catamarans comparing Waterjet and Screw Propeller Performance using Model Testing
  author:
  - family: Kamal
    given: Iwan Mustaffa
  - family: Zürcher
    given: Konrad
  - family: Bose
    given: Neil
  - family: Binns
    given: Jonathan
  - family: Chai
    given: Shuhong
  - family: Davidson
    given: Gary
  URL: 'http://www.marinepropulsors.com/proceedings/2015/MA3-2.pdf'
  publisher: Fourth International Symposium on Marine Propulsors
  issued:
    year: 2015
    month: 6

- id: lu2015
  title: CFD prediction and simulation of a pumpjet propulsor
  author:
  - family: Lu
    given: Lin
  - family: Pan
    given: Guang
  - family: Sahoo
    given: Prasanta K.
  URL: 'http://www.sciencedirect.com/science/article/pii/S2092678216000042'
  publisher: International Journal of Naval Architecture and Ocean Engineering
  issued:
    year: 2015
    month: 10

- id: giles2010
  title: The Advanced WaterJet Propulsor Performance and Effect on Ship Design
  author:
  - family: Giles
    given: Will
  - family: Dinham-Peren
    given: Tom
  - family: Amaratunga
    given: Shane
  - family: Vrijdag
    given: Arthur
  - family: Partridge
    given: Richard
  URL: 'https://www.bmtdsl.co.uk/media/6097875/BMTDSL-The-Advanced-Waterjet-Confpaper-INEC-May10.pdf'
  publisher: Tenth International Naval Engineering Conference
  issued:
    year: 2010
    month: 5

- id: haimov2010
  title: Ducted Propellers. A Solution for Better Propulsion of Ships. Calculations and Practice.
  author:
  - family: Haimov
    given: H.
  - family: Bobo
    given: M.J.
  - family: Vicario
    given: J.
  - family: del Corral
    given: J.
  URL: 'http://www.cehipar.es/__files/users/publicaciones/217.pdf'
  publisher: Ministerio de Defensa
  issued:
    year: 2010
    month: 5

- id: abdel2010
  title: Design of a Multi-Component Propulsor
  author:
  - family: Abdel-Maksoud
    given: Moustafa
  - family: Steden
    given: Max
  - family: Hundemer
    given: Jochen
  publisher: 28th Symposium of Naval Hydrodynamics
  issued:
    year: 2010
    month: 9

- id: bakker2006
  title: Boundary Layers and Separation
  author:
  - family: Bakker
  URL: 'http://www.bakker.org/dartmouth06/engs150/11-bl.pdf'
  publisher: Dartmouth College
  issued:
    year: 2002

- id: mcbride1979
  title: Design and analysis of turbomachinery in an incompressible, steady flow using the streamline curvature method
  author:
  - family: McBride
    given: Mark W.
  URL: 'www.dtic.mil/dtic/tr/fulltext/u2/a071107.pdf'
  publisher: Naval Sea Systems Command
  issued:
    year: 1979
    month: 7

- id: wislicenus1986
  title: Preliminary Design of Turbopumps and Related Machinery
  author:
  - family: Wislicenus
    given: George F.
  URL: 'https://ntrs.nasa.gov/archive/nasa/casi.ntrs.nasa.gov/19870008232.pdf'
  publisher: NASA
  issued:
    year: 1986
    month: 10

- id: wislicenus1973
  title: Hydrodynamic Design Principles of Pumps and Ducting for Waterjet Propulsion
  author:
  - family: Wislicenus
    given: George F.
  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/775620.pdf'
  publisher: Naval Ship Research & Development Center
  issued:
    year: 1973
    month: 6

- id: brandau1967
  title: Apects of performance evaluation of waterjet propulsion systems and a critical review of the state-of-the-art
  author:
  - family: Brandau
    given: John H.
  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/827069.pdf'
  publisher: Naval Ship Research & Development Center
  issued:
    year: 1967
    month: 10

- id: henderson1964
  title: A Method for the Design of Pumpjets
  author:
  - family: Henderson
    given: R.E.
  - family: McMahon
    given: J.F.
  - family: Wislicenus
    given: G.F.
  URL: 'http://www.dtic.mil/dtic/tr/fulltext/u2/439631.pdf'
  publisher: Ordinance Research Laboratory, Department of Navy
  issued:
    year: 1964
    month: 5

- id: MAN2017
  title: Basic Principals of Ship Propulsion
  author:
  - family: MAN Diesel & Turbo
  URL: 'https://marine.man.eu/docs/librariesprovider6/propeller-aftship/basic-principles-of-propulsion.pdf?sfvrsn=0'
  publisher: MAN Diesel & Turbo
  issued:
    year: 2017

- id: oosterveld1970
  title: Wake Adapted Ducted Propellers
  author:
  - family: Oosterveld
    given: IR.M.W.C.
  URL: 'http://www.marin.nl/web/Publications/Publication-items/Wake-Adapted-Ducted-Propellers.htm'
  publisher: Netherlands Ship Model Basin Wageningen - Netherlands
  issued:
    year: 1970

- id: davies2017
  title: Pump up the pump jet?
  author:
  - family: Davies
    given: Andrew
  URL: 'https://www.aspistrategist.org.au/pump-up-the-pump-jet/'
  publisher: Australian Strategic Policy Institute
  issued:
    year: 2017
    month: 11

- id: stanford2017
  title: Australia's Future Submarine Getting This Key Capability Right
  author:
  - family: Jon
    given: Stanford
  URL: 'http://www.insighteconomics.com.au/reports/2017_Insight_Economics_Submarine_Report.pdf'
  publisher: Insight Economics
  issued:
    year: 2017
    month: 09

  - id: willemsen2013
  title: Improving Potential Flow Predictions for Ducted Propellers
  author:
  - family: J.C
    given: Willemsen
  URL: 'http://www.refresco.org/wp-content/uploads/2015/03/2013-Msc_Thesis_ChrisWillemsen.pdff'
  publisher: Marin
  issued:
    year: 2013
    month: 12

---

@@

# Contents

# Introduction

The purpose of this paper is to investigate the suitability of the pump-jet selection for Australia's future fleet of submarines.  In particular, the question of the relative efficiency, and quietness of pump-jets in comparison to open propellers is of particular importance.  Much has been made of the significance of the pump-jet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pump-jet rendered propellers obsolete.

Necessarily, this topic depends upon a considerable amount of physics, particularly around hydrodynamics, fluid mechanics, and turbomachinery, all of which might be difficult or daunting topics to usefully inform a public policy debate. However, in the case of a $50 billion dollar military acquisition, neglecting to engage with the technical matters that are so crucial to such a decision would be deeply foolish.  To that end, this paper has a very specific ambition.  It doesn't seek to advance the technical field with any new research or insight.  Instead, it aims to make as much of the relvant physics as possible to this particular question comprehensible in laymans terms, and connect the essential concepts as directly as possible with the most important conclusions of public import.

As such, it will tend to make frequent use of diagrams, illustrations, as well as simple prose, and relatively infrequently use mathematical equations, (unless of crucial importance), though some effort is made to point to references where the fuller and more formal derivations of these relationships can be found.

It should be noted that the technical fields involved here are vast and rapidly evolving.  The author makes no claim to be a world authority on the entirety of the topics which are touched. It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.

However, the nature of scientific discovery and technical progress is that many of best discoveries are actually built-upon and confirmed, rather than swept aside, by later developments.  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.  Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints that will be inherent in that field.

In this paper, since we cannot know the precise details of the particular state of the art in largely classified military programs, I will attempt to be particularly clear about those things for which we can have great confidence (often broad principals or relationships which determine trade-offs) and those things which we can't (the precise degrees or points and sensitivities).  To that end, in this paper I've underaken modelling based upon the broad principals which we can have confidence, and conducted sensitivity testing based on a range of plausible values which seem realistic for those things about which we can have less confidence. I have also, in discussion with experts and through reading the available literature, attempted where possible to identify the most plausible of the uncertain values, in order to advance discussion.

With the support of the principal sponsor of this paper, the relevant models may be found in a user-friendly format as a web-app to allow their further interrogation and testing for other plausible scenarios.  In addition, for increased transparency, the relevant code and equations underlying are available on github.

# Executive Summary

#### The difference between nuclear and conventional for speed
This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine. A crucial input to the investigation is a rough understanding of what the speeds of operation are likely to be for a submarine.  Whereas nuclear submarines are reportedly capable of reaching speeds in excess of 30kt, and might transit long distances at such high speeds, conventional sumbarines are not thought of as being able to reach speeds far above 20kt in a sprint, and can only sustain speeds of 8-10kt for long-distance transits.  Moreover, on patrol, a large portion of thier work is done at very low speeds, typically thought to be in the range of 2-4kt.

#### The importance of low speed operation for diesel-electric submarines
The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.  Consequently, excessive energy consumption results in greatly reduced dived endurances.  Since a submarine's position is vastly more likely to be discovered when it is on the surface operating its diesel engines, this ability to remain submerged for a long time is crucially important for combat operations, and in transiting through sensitive or contested areas.  For nuclear submarines, which posess a practically infinite supply of energy from the nuclear reactor, efficiency at low speeds is of no concern.  In fact, dispersing additional energy, (provided it can be done quietly) is probably advantageous for a nuclear submarine, since it will allow the nuclear reactor to avoid running at very low power levels, where the stability of the reactor is reduced.

#### What is cavitation
Pumpjets have been widely adopted by navies operating nuclear submarines.  The principal advantage of a pumpjet relevant to submarines is their ability to avoid problems associated with cavitation, which is known to occur for propellers attempting to operate at high speeds.  Cavitation, which is the rapid expansion and collapse of a bubble or void in the water, is particularly problematic for submarines, since it results in the creation of a great deal of noise which could be detected by an enemy.

#### How does a pump-jet work
A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.  This means that the blades can turn at a lower speed relative to the water they are connected with, allowing a less violent action, which induces less cavitation.  In this way, the jet does it's work less by directly acelerating the water, but by raising its pressure.  This raised pressure is converted back to movement, which produces thrust, as it exits the jet and returns to the same pressure as the surrounding environment.

#### The necessity of drag, and decelleration, induced by the duct
The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be raised around the impeller, in a way that is not possible for an open propeller.  By a fundamental requirements of physics, this actually requires that the water's incoming speed be **reduced** before it reaches the impeller. (The kinetic energy embodied in movement is converted into potential energy, or pressure.)  Whilst the duct narrowing at the nozzle also necessarily accelerates the water, (as the additinal pressure imparted by the impeller is converted back to kinetic energy) it's an essential feature of all pump-jets that the water flow is decelerated at the point of reaching the impeller. Consequently, an elementary form of the pump-jet is also termed the 'decelerating duct' applied to a propeller.  Put simply, the water has to slow down to go fast again.

#### What happens to a jet at low speed, in simple terms
The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.  This is the case when a jet designed for high-speed attempt to operate at a dramatically reduced speed.  (Not to be confused with a vessel with very little water speed working very hard, as might be the case for a tug or barge.)  In this case, the slowing down and speeding up results in an unnecssary additional step which reduces greatly the efficiency of propulsive system. Or, put another way, overcomign the resistance of moving water through the shroud becomes much greater relative to the total thrus produced by the jet.

#### Why this isn't easily noticeable in normal circumstances
It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).  This doesn't mean that they don't still work at low speeds and produce some thrust.  All it means is that far more power per unit of thrust will be required than might be at other speeds, as a higher fraction of energy is expended producing the turblence (random, round-and-round movement) inside and around the jet shroud than goes towards direct front-to-back acceleration of the water column, which produces thrust.

#### What the submarine requirement tells us about the duct
Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.  Moreover, the particular demands of a pump-jet suited to a nuclear submarine, (eliminating all cavitation in as wide a range of operating circumstances as possible) considerably narrows the plausible range concerned.  In order to minimise cavitation, the degree of pressure elvation at the impeller would be relatively high, or the total surfaces of the impeller blades much aslo become larger.  Both of these design requirements necessitate changes to the duct or blade designs which would be in tension with overall efficiency, and most pronounced at the lowest of operating speeds.

#### Quantitative Conclusions
In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine. The most plausible scenarios I find include the reduction of dived endrance and range between 20% and 50%, or effectively halving time and distance that submarine may remain submerged for during combat operations at speeds around 3-5kt.

#### The acoustic advantage of jets at higher speed
My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation. Indeed, it has been remarked by naval researchers that pump-jets could be designed which would not cavitate past the point when the body they propel experienced caviation.   As such, it seems perfectly plausible that pump-jets confer one advantage on a submarine, in that they can accelerate to a higher speed without cavitation occurring.  This is known as a higher 'tactical silent speed'.

#### Turbulence and flow separation at low speed
However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.  Moreover, a jet will necessarily incur substantially larger degrees and types of turbulence in order to produce net thrust in this regime.  These likely include discontinuities and instability in the flow entering the duct and passing through the impeller and stator (flow separation) as the water is acelerated sufficient to be slowed, and then re-accelerated.  In certain circumstances where resonances might arise these could be highly adverse to acoustic performance.  However, assuming that by careful design such resonances can be eliminated (which we assume has been achieved) these effects would not result in any cavitation, and would only increase the noise attributable to turbulence in solid water, which is far less than for any cavitation.

#### The acoustic question at low speed
But in either case, given the necessarily raised levels of turblence generated by a jet than a propeller at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.  These could be substantial in a directions perpendicular to the direction of travel, but would be much smaller when viewed from the aft or forward directions.  It should also be noted that many underwater acoustic environments, sound tennds to reflect and bend in different directions as it propagates. As such, the claim that pump-jets are universally quieter than propellers should be treated with caution.  In a variety of circumstances, including most operations at patrol speed, it may not always be true.

#### The inevitable trade-off
However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.  These are a lower dived endurance, a lower dived range, a lower overall endurance, a lower overall range, and a worse indescretion ratio.

#### The contradiction between stated requirement and chosen technology
Given the substantial emphasis which was placed on overall range and endurance of the submarine, it is difficult to understand how one particular unique requirement would be elevated so high above the other strategic and tactical advantages afforded by propellers.

# Speed and Drag - Why very slow is very very $(very)^2$ economical

Perhaps the most important relationship to understand is the relationship between speed and drag as it pertains to submarines.  This is immportant because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.  (It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)

Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it. Quite simply, it's a force that acts in the opposite direction.  There are multiple sources of drag for different types of scenarios.  For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies. For a fuller discussion of types of drag, see [@carlton2007]  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.  The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.  For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.  The drag is also directly proportional to the density of the fluid being moved through, $\rho$.  (Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)

But the most sensitive factor in this relationship is the speed at which the object moves through the liquid.  The drag increases not with the speed ($v$), but with the **square** of the speed . This means if the speed doubles, the drag increases by four.  If the speed triples, the drag increases by a factor of nine.

$$F_D = \frac{1}{2}\rho v^2C_DA$$

(Technical aside: This law applies wherever the the flow over the surface is turbulent. It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case wher Reynolds numbers are less than 1.  Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.  Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull).

It's crucial to understand, however, that drag is only a force, and doesn't directly inform us about how much energy is consumed, until we multiply it by the **distance** over which it is applied, not the time for which it is applied.  To think about it simply, gravity exerts a force on you downwards all the time.  But you don't expend any energy overcoming it when sitting still.  If you climb stairs, the amount of gravitational potential you attain depends on how high up you climb, not how long you spend on the ladder or stair-case.

This has a significant consequence for propulsion, since the amount of power (energy expended in a given time) required for thrust scales with the drag, multiplied by the distance covered in a given time.  As such, power required (and fuel/battery consumption) scales with velocity **cubed** rather than velocity squared.  This means that if you double your speed of travel, fuel consumption for a given period of time will increase by a factor of eight.  Due to the increased speed of travel, the fuel required to cover a given distance will only quadruple.  Hence the range that can be covered with a given amount of energy (assuming propulsion dominates energy requirements) tends to scale with the inverse of velocity squared, whereas the endurance (amount of time that can be spent travelling) scales with the inverse of velocity cubed.

This has a profound impact on the operation and engineering of maritime vessels.  Reducing velocity has such a substantial reduction on hull drag that going slower is almost always a reasonable means of conserving fuel overall.  To the extent that time is non-critical, slower is always much much better.  It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming [@liang2014] in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods [@sanguri2012 pp. 8-10]. The fuel saving from operating even 30% slower means that around half the total energy is required for a given journey.  Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.

This effect is of profound importance to understanding the operations of conventional submarines, which have extremely constrained energy stores when operating under the surface. Diesel fuel when burned has an energy density of approximately 45MJ/kg.  In contrast, a lead-acid battery might have energy densities in the range of 0.08-0.14MJ/kg. With something like 400 times as much energy per kg embarked in diesel, operating on batteries imposes an extreme demand for economy on propulsive power.  Happily for submarines, the square law for drag, and cubed law for power, allow an almost commensurate reduction in power demands to take place by slowing down to very slow speeds when submerged. Hypothetically, a conventional submarine might have a total range of 10,000nm from using it's diesel payload at 8kt.  If an equal weight of lead-acid batteries as fuel were carried, the submarine could only travel about 25nm when submerged at the same speed in a single charge.  But by travelling at 4kt, that quickly increases to 100nm, and 400nm at 2kt (neglecting hotel load here for simplicity).

It's worthwhile pointing out that extreme demand for economy imposed by the poor energy density of batteries distinguishes submarines quite remarkably from other types of boat or ship design.  In every other application, the propulsion system is designed around a particular speed and loading condition which it is optimally efficient for, and a band of plausible variation around this.  For example, a cargo ship designed for operation at 22kt will have it's propeller and engine (propulsion system) very carefully designed around these speeds and the plausible levels of loading at those speeds. A top-speed which might be somewhat higher than 22kt would be calculated, but would be relatively unlikely to be used for any purpose other than an emergency.  Lower speeds might be considered 15-18kt for the purposes of slow-steaming.  However, the relative efficiency at 3kt is of little concern, since the drag at that speed will be under 2% of what it would be at 22kt, and the power required around 0.25%.  Even if the efficiency was substantially worse, or better (by a factor of two or three even) the impact on the overall economics of the operation would be negligible when compared to marginal improvements at the design speed.

The same consideration applies to other types of vessels, which might use water-jets.  Fast pilot-boats, for instance, might spend quite a considerable time manouvering in and out of harbour or alongside at very low waterspeeds. However, their overall fuel consumption is still likely to be dominated by the fast section of their trip to a ship.  If the in-harbour water-speed was 8kt, but 32kt was the optimum speed for the open-water section of their journey, the fuel requirement for a given distance could still be over ten times greater at 32kt. (Aside, the physics described above would say 16 times, but some reduction due to the reduction of surface area from planing might offset this.)  Consequently the owner/operator would be at least ten times as concerned about efficiency at 32kt as at 8kt, (unless the harbour transit was much much longer than the open water section), and have negligible concern about effiencies at 2kt or 3kt, which would rarely amount to more than a couple of percent of total propulsion energy demand.

>>>>

@@

## Executive Summary

This paper was commissioned in order to investigate whether pumpjets could plausibly be as efficient, or more efficient than a suitably designed open propeller for a conventional submarine. A crucial input to the investigation is a rough understanding of what the speeds of operation are likely to be for a submarine.  Whereas nuclear submarines are reportedly capable of reaching speeds in excess of 30kt, and might transit long distances at such high speeds, conventional sumbarines are not thought of as being able to reach speeds far above 20kt in a sprint, and can only sustain speeds of 8-10kt for long-distance transits.  Moreover, on patrol, a large portion of thier work is done at very low speeds, typically thought to be in the range of 2-4kt.

The efficiency of the propulsion system at such low speeds is of great significance for a conventional submarine, since it must rely on batteries or other air-independent propulsion sources for power when entirely submerged.  Consequently, excessive energy consumption results in greatly reduced dived endurances.  Since a submarine's position is vastly more likely to be discovered when it is on the surface operating its diesel engines, this ability to remain submerged for a long time is crucially important for combat operations, and in transiting through sensitive or contested areas.  For nuclear submarines, which posess a practically infinite supply of energy from the nuclear reactor, efficiency at low speeds is of no concern.  In fact, dispersing additional energy, (provided it can be done quietly) is probably advantageous for a nuclear submarine, since it will allow the nuclear reactor to avoid running at very low power levels, where the stability of the reactor is reduced.

Pumpjets have been widely adopted by navies operating nuclear submarines.  The principal advantage of a pumpjet relevant to submarines is their ability to avoid problems associated with cavitation, which is known to occur for propellers attempting to operate at high speeds.  Cavitation, which is the rapid expansion and collapse of a bubble or void in the water, is particularly problematic for submarines, since it results in the creation of a great deal of noise which could be detected by an enemy.

A fundamental requirement of a pump-jet in order to avoid this cavitation is that the the working parts of the jet (the rotating blades inside it, known as the impeller) operate at a higher pressure than propeller blades operating in open water would.  This means that the blades can turn at a lower speed relative to the water they are connected with, allowing a less violent action, which induces less cavitation.  In this way, the jet does it's work less by directly acelerating the water, but by raising its pressure.  This raised pressure is converted back to movement, which produces thrust, as it exits the jet and returns to the same pressure as the surrounding environment.

The role of the shroud (or tunnel, mantel, duct) around the jet is to allow the pressure to be so raised around the impeller, in a way that is not possible for an open propeller.  By a fundamental requirements of physics, this actually requires that the water be **reduced** before it reaches the impeller. (The kinetic energy embodied in movement is converted into potential energy, or pressure.)  Whilst the duct narrowing at the nozzle also necessarily accelerates the water, (as the additinal pressure imparted by the impeller is converted back to kinetic energy) it's an essential feature of all pump-jets that the water flow is decelerated at the point of reaching the impeller. Consequently, an elementary form of the pump-jet is also termed the 'decelerating duct' applied to a propeller.  Put simply, the water has to slow down to go fast again.

The problems for pump-jets arise when the water is already going slow, and you don't want it to go that much faster.  This is the case when a jet designed for high-speed attempt to operate at a dramatically reduced speed.  (Not to be confused with a vessel with very little water speed working very hard, as might be the case for a tug or barge.)  In this case, the slowing down and speeding up results in an unnecssary additional step which reduces greatly the efficiency of propulsive system. Or, put another way, overcomign the resistance of moving water through the shroud becomes much greater relative to the total thrus produced by the jet.

It is for this reason that waterjets of any kind are known to have an efficiency curve that falls off towards zero as the net thrust they produce also diminishes to zero (which for a given vessel will correspond to water speed).  This doesn't mean that they don't still work at low speeds and produce some thrust.  All it means is that far more power per unit of thrust will be required than might be at other speeds, as a higher fraction of energy is expended producing the turblence (random, round-and-round movement) inside and around the jet shroud than goes towards direct front-to-back acceleration of the water column, which produces thrust.

Whist estimating the exact shape and level of the efficiency curve for a particular pump-jet and propeller is impossible without detailed knowledge of their design, the over all trends of their shapes in the extremes can be known from well-established principals.  Moreover, the particular demands of a pump-jet suited to a nuclear submarine, (eliminating all cavitation in as wide a range of operating circumstances as possible) considerably narrows the plausible range concerned.  In order to minimise cavitation, the degree of pressure elvation at the impeller would be relatively high, or the total surfaces of the impeller blades much aslo become larger.  Both of these design requirements necessitate changes to the duct or blade designs which would be in tension with overall efficiency, and most pronounced at the lowest of operating speeds.

In the production of this paper I have developed a computational model which maps the impact of different efficiency curves directly to the dived range and endurance of a submarine. The most plausible scenarios I find include the reduction of dived endrance and range between 20% and 50%, or effectively halving time and distance that submarine may remain submerged for during combat operations at speeds around 3-5kt.

My review of the literature confirms that a pumpjet may produce a significant acoustic advantage in circumstances where an open propeller would experience any degree of cavitation. Indeed, it has been remarked by naval researchers that pump-jets could be designed which would not cavitate past the point when the body they propel experienced caviation.   As such, it seems perfectly plausible that pump-jets confer one advantage on a submarine, in that they can accelerate to a higher speed without cavitation occurring.  This is known as a higher 'tactical silent speed'.

However, at much lower speeds such as patrol speeds (2-4kt) it is most likely that a propeller will be able to operate well below the point of any cavitation inception, and would consequently also be extremely quiet.  Moreover, a jet will necessarily incur substantially larger degrees and types of turbulence in order to produce net thrust in this regime.  These likely include discontinuities and instability in the flow entering the duct and passing through the impeller and stator (flow seperation) as the water is acelerated sufficient to be slowed, and then re-accelerated.  In certain circumstances where resonances might arise these could be highly adverse to acoustic performance.  However, assuming that by careful design such resonances can be eliminated (which we assume has been achieved) these effects would not result in any cavitation, and would only increase the noise attributable to turbulence in solid water, which is far less than for any cavitation.

But in either case, given the necessarily raised levels of turblence generated by a jet at very low speeds, the claim that the jet is quieter in this regime must rely entirely upon shielding effects from the shroud.  These could be substantial in a directions perpendicular to the direction of travel, but would be much smaller when viewed from the aft or forward directions.  It should also be noted that many underwater acoustic environments, sound tennds to reflect and bend in different directions as it propagates. As such, the claim that pump-jets are universally quieter than propellers should be

However, as a direct and necessary result of this higher tactical silent speed, some disadvantages will be incurred, owing to the additional drag induced by the shroud which prevents cavitation at the impeller.  These are a lower dived endurance, a lower dived range, a lower overall endurance, a lower overall range, and a worse indescretion ratio.

Given that

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
